How does my active transport access change when car access is restricted? Measuring multi-modal accessibility with spatial availability

### 2023-05-30 Multi-mode-acc-methods.qmd
draft of intro and illustrative example - for AP review.

---
title: "How does my active transport access change when car access is restricted? Measuring multi-modal accessibility with spatial availability"
format: docx
bibliography: bibliography.bib
---

@@

<!-- examples of constrained (competitive) measures, the introduction of spatial availability -->

The third type of accessibility measure, competitive measures, discount the number of opportunities allocated based on the population that is demanding those opportunities in addition to discounting by travel impedance (like unconstrained accessibility measures). Competitive measures hence consider both the supply of the opportunities and the demand for those opportunities by the population in addition to the travel impedance. They have been criticized for being difficult to compute (relative to unconstrained measures) and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020]. In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of competitive accessibility measure that is _constrained_ beside a popular _unconstrained_ accessibility measure (Hansen-type gravity model) and a competitive accessibility measure (two-step floating catchment area (2SFCA)). We title spatial availability a _constrained_ accessibility measure since it proportionally allocates all the opportunities (supply) to each origin-location based on competition from the population (demand) and travel impedance in a single step with an interpretable output. The output is the number of opportunities available in the region, per location, such that if all the values for the locations in the region were summed they would equal the total opportunities in the region.

@@

However, this is a gap in the literature. Researchers and transport planners needs accessibility tools that interpretable consider competition. This is the aim of this paper. An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility. Take for instance, measuring the accessibility to opportunities for a region for car and transit modes. The literature suggests that these modes are in direct competition. A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other. Thus, the potential for the population to interact with opportunities is directly dependant on ... the opportunities..

@@

A clustering method to modify the cumulative accessibility measure @mohriClusteringMethodMeasuring2021
<!-- summary of the paper-->
With the inconsistencies and limitations of existing measures in mind, we propose the multimodal application of spatial availability. To illustrate this application, in Section 2, we demonstrate the formulation of spatial availability in a simple example of two modes and three population and opportunity centers in a synthetic example. In Section 3, we then demonstrate an empirical example of the low emission zone in Madrid Spain. Specifically, how an expansion in the restriction of private vehicles changes the access to opportunities for transit, cycling and walking. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, potential future uses in policy planning scenarios, and limitations of accessibility measures.

@@

In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities $O$ that are allocated to $i$ from all destinations $j$:
$$
V_i = \sum_{j=1}^NO_jF^t_{ij}
$$

@@

-   $F^t_{ij}$ is a balancing factor that depends on the population and cost of movement in the system.
-   $O_j$ is the number of opportunities at $j$.
-   $V_i$ is the number of spatially available opportunities from the perspective of $i$.

The general form of spatial availability is also as a sum, and the fundamental difference between Hansen- and Shen-type accessibility is that opportunities are allocated proportionally. Balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).

We adapt the illustrative example used in @shen1998 assuming two modes and demonstrate spatial availability's within this multimodal application. <!-- TO DO: demonstrate an illustrate of the example-->

@@

According to the gravity modelling framework, the potential for interaction depends on the mass (i.e., the population) and the friction of distance (i.e., the impedance function). We begin by describing the proposed proportional allocation mechanism based on demand by the population. Recall, the total population in the example is 210,000. The proportion of the population by population center, by the two modes ($x$, and $z$) is:
$$
\begin{array}{l}
F^p_A = \frac{50,000}{210,000} & \text{where } F^{px}_A = \frac{40,000}{210,000} \text{ & } F^{pz}_A = \frac{10,000}{210,000}\\ 
\\
F^p_B = \frac{150,000}{210,000} & \text{where } F^{px}_B = \frac{100,000}{210,000} \text{ & } F^{pz}_B = \frac{50,000}{210,000}\\
\\
F^p_C = \frac{10,000}{210,000} & \text{where } F^{px}_C = \frac{7,000}{210,000} \text{ & } F^{pz}_C = \frac{3,000}{210,000}\\
\end{array}
$$

Competition between the subsets of the populations that use the jobs matters. After all, there is a finite number of jobs (210,000 in our example). Jobs are allocated proportionally from each employment center to each population center depending on their population sizes as per the balancing factors $F^p_i$. In this way, employment center 1 allocates $100,000\cdot \frac{50,000}{210,000}= 23,809.52$ jobs to $A$. This  $100,000\cdot \frac{150,000}{210,000}= 71,428.57$ jobs to $B$; and $100,000\cdot \frac{10,000}{210,000}= 7,142.857$ jobs to $C$. Notice how this mechanism ensures that the total number of jobs at employment center 1 is preserved at 100,000.

We can verify that the number of jobs allocated is consistent with the total number of jobs in the system:
$$
\begin{array}{l}
\text{Employment center 1 to population centers A, B, and C: }\\
100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
\\
\text{Employment center 2 to population centers A, B, and C: }\\
100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
\\
\text{Employment center 3 to population centers A, B, and C: }\\
10,000 \cdot \frac{50,000}{210,000} + 10,000 \cdot \frac{150,000}{210,000} + 10,000 \cdot \frac{10,000}{210,000} = 10,000\\
\end{array}
$$

In the general case where there are $N$ population centers in the region, we define the following population-based balancing factors in Equation (\ref{eq:population-balancing-factor}):
```{=tex}
\begin{equation}
\label{eq:population-balancing-factor}
F^{pM}_{i} = \frac{P_{i}^m}{\sum_{i=1}^N P_{i}^m}
\end{equation}
```

Balancing factor $F^p_{i}$ corresponds to the proportion of the population in origin $i$ relative to the population in the region. On the right hand side of the equation, the numerator $P_{i}^m$ is the population at origin $i$. <!-- The summation in the denominator is over $i=1,\cdots,N$, and adds up to the total population of the region. Notice that we incorporate an empirical parameter $\alpha$. The role of $\alpha$ is to modulate the effect of demand by population. When $\alpha <1$, opportunities are allocated more rapidly to smaller centers relative to larger ones; $\alpha>1$ achieves the opposite effect.-->

The discussion so far concerns only the mass effect (i.e., population size) of the gravity model. In addition, the potential for interaction is thought to decrease with increasing cost, so next we define similar balancing factors but based on the impedance.

@@

Clearly, using only balancing factor $F^p_{i}$ to calculate spatial availability $V^p_i$ does not account for the cost of reaching employment centers. Consider instead a set of balancing factors $F^c_{ij}$ that account for the friction of distance for our example. Recall that the impedance function $f(c_{ij})$ equals $\exp(-\beta\cdot c_{ij})$ where  $\beta = 0.1$ and travel time $c_{ij}$ is either 15, 30 or 60 minutes. For instance, the impedance-based balancing factors $F^c_{ij}$ would be the following for employment center 1 (employment center 2 and 3 have their own balancing factor values for each origin $i$ as will be discussed later):
$$
\begin{array}{l}
F^c_{A1} = \frac{0.223130}{0.223130 + 0.049787 + 0.000045} = 0.8174398\\
F^c_{B1} = \frac{0.049787}{0.223130 + 0.049787 + 0.000045} = 0.1823954\\
F^c_{C1} = \frac{0.000045}{0.223130 + 0.049787 + 0.000045} = 0.0001648581\\
\end{array}
$$

Balancing factors $F^c_{ij}$ use the impedance function to proportionally allocate more jobs to closer population centers, that is, to those with populations _more willing to reach the jobs_. Indeed, the factors $F^c_{ij}$ can be thought of as the proportion of the population at $i$ willing to travel to destination $j$, conditional on the travel behavior as given by the impedance function. For instance, ${81.74398}\%$ of jobs from employment center 1 are allocated to population center $A$ based on impedance.

So as follows from our example, of the 100,000 jobs at employment center 1 the number of jobs allocated to population center $A$ is $100,000\times 0.8174398 = 81,743.98$ jobs; the number allocated to population center $B$ is $100,000\times 0.1823954 = 18,239.54$ jobs; and the number allocated to population center $C$ is $100,000\times 0.0001648581 = 16.48581$ jobs. We see once more that the total number of jobs at the employment center is preserved at 100,000. In this example, the proportional allocation mechanism assigns the largest share of jobs to population center $A$, which is the closest to employment center 1 and the smallest to the more distant population center $C$.

In the general case where there are $N$ population centers and $J$ employment centers in the region, we define the following impedance-based balancing factors:
```{=tex}
\begin{equation}
\label{eq:impedance-balancing-factor}
F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^N f(c_{ij})}
\end{equation}
```

The total number of jobs available to $i$ from $j$ according to impedance is defined as follows:
$$
V^c_{ij} = O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
$$

The total number of jobs available to $i$ from all destinations is:
$$
V^c_{i} = \sum_{j=1}^J O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
$$

>>>>

---
title: "How does my active transport access change when car access is restricted? Measuring multi-modal accessibility with spatial availability"
format: docx
bibliography: bibliography.bib
---
```{r load-packages, include=FALSE, cache=FALSE}
# library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax

# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# options(scipen = 999)

library(patchwork) # The Composer of Plots
library(sf) # Simple Features for R
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggforce) # Accelerating 'ggplot2'
library(geomtextpath) # Curved Text in 'ggplot2'
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
library(scales) # Scale Functions for Visualization
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function
sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  # Sum of population in system
  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  # Balancing factor size
  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  # Sum of impedance
  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  # Join sum of impedance to table
  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  # Balancing factor impedance
  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  # Add balancing factors to table
  x$f_c <- f_c
  x$f_p <- f_p

  # Sum of balancing factors for mass and impedance
  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))

  # Joint balancing factor
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  dplyr::pull(x, !!opp) * f_t
}

#detailed
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  x$f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * f_t)
}
```

```{r synthetic-data, include=FALSE}
od_tt <- data.frame(i = c("A", "A", "A", "B", "B", "B", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", "1", "2", "3"), # Three destinations
                    tt_z = c(15, 30, 100, 30, 15, 100, 100, 100, 15), # Travel time, z-mode (original)
                    tt_x = c(10, 25, 80, 25, 10, 80, 80, 80, 10), # Travel time, x-mode (the faster mode - like car)
                    pop = c(50000, 50000, 50000, 150000, 150000, 150000, 10000, 10000, 10000), # Population
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000, 100000, 100000, 10000)) # Jobs

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("z", "z", "z", "x", "x", "x",
                          "z", "z", "z", "x", "x", "x",
                          "z", "z", "z", "x", "x", "x"),
                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
                           30, 15, 100, 25,  10,  80,
                           100, 100, 15, 80, 80, 10), 
                    pop = c(10000, 10000, 10000, 40000, 40000, 40000,
                            50000, 50000,50000, 100000, 100000, 100000,
                            3000, 3000, 3000, 7000,7000, 7000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```

```{r data-figure-with-toy-example, include=FALSE}
od <- data.frame(id = c("A", "B", "C", "1", "2", "3"),
                 type = c("Population", "Population", "Population", "Jobs", "Jobs", "Jobs"),
                 size = c(50000, 150000, 10000, 100000, 100000, 10000),
                 x = c(2.5, 2.5, 6.5, 0.5, 0.5, 6.5),
                 y = c(8.0, 2.5, 4.5, 7.5, 2.5, 2.5))

centers <- data.frame(id = c("Urban center", "Suburb", "Satellite town"),
                 radius = rep(1.3, 3),
                 x = c(1.5, 1.5, 6.5),
                 y = c(2.5, 7.5, 3.5))

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (z)\n10 min (x)", "30 min (z)\n25 min (x)", "100 min (z)\n80 min (x)",
                                "30 min (z)\n25 min (x)", "15 min (z)\n10 min (x)", "100 min (z)\n80 min (x)",
                                "100 min (z)\n80 min (x)", "100 min (z)\n80 min (x)", "15 min (z)\n10 min (x)"))

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop", "150,000 pop", "10,000 pop", "100,000 jobs", "100,000 jobs", "10,000 jobs"))
```

```{r, creating-toy-example-plot, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=3) +
  # Create table
  annotate(geom = "table",
           x = 8, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("images/Fig1.png",dpi=600)
```

@@

<!-- examples of constrained (competitive) measures, the introduction of spatial availability -->
The third type of accessibility measure, competitive measures, discount the number of opportunities allocated based on the population that is demanding those opportunities in addition to discounting by travel impedance (like unconstrained accessibility measures). Competitive measures hence consider both the supply of the opportunities and the demand for those opportunities by the population in addition to the travel impedance. They have been criticized for being difficult to compute (relative to unconstrained measures) and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020]. In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of competitive accessibility measure that is _constrained_ beside a popular _unconstrained_ accessibility measure (Hansen-type gravity model) and a competitive accessibility measure (two-step floating catchment area (2SFCA)). We title spatial availability a _constrained_ accessibility measure since it proportionally allocates all the opportunities (supply) to each origin-location based on competition from the population (demand) and travel impedance in a single step with an interpretable output. The output is the number of opportunities available in the region, per location, such that if all the values for the locations in the region were summed they would equal the total opportunities in the region.

ANother example of proportional allocation, BFCA

@@

However, this is a gap in the literature. Researchers and transport planners needs accessibility tools that interpretable consider competition. This is the aim of this paper. We plan to expland to this measure to multi-modal systems. opportunities are avialble to users of different modes. The cost of reaching them is not uniform.

<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility. Take for instance, measuring the accessibility to opportunities for a region for car and transit modes. The literature suggests that these modes are in direct competition. A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other. Thus, the potential for the population to interact with opportunities is directly dependant on ... the opportunities..  -->

@@

A clustering method to modify the cumulative accessibility measure @mohriClusteringMethodMeasuring2021

<!-- summary of the paper-->
With the inconsistencies and limitations of existing measures in mind, we propose the multimodal application of spatial availability. To illustrate this application, in Section 2, we demonstrate the formulation of spatial availability in a simple example of two modes and three population and opportunity centers in a synthetic example. In Section 3, we then demonstrate an empirical example of the low emission zone in Madrid Spain. Specifically, how an expansion in the restriction of private vehicles changes the access to opportunities for transit, cycling and walking. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, potential future uses in policy planning scenarios, and limitations of accessibility measures.

@@

In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities $O$ that are allocated to $i$ from all destinations $j$:
$$
V_i = \sum_{j=1}^J O_jF^t_{ij}
$$

@@

-   $V_i$ is the number of spatially available opportunities at $i$.
-   $O_j$ is the number of opportunities at $j$.
-   $F^t_{ij}$ is a balancing factor that depends on the demand for opportunities and cost of movement in the system.

The general form of spatial availability is a sum: this is similar to Hansen-type accessibility (popular _unconstrained_ measure) and Shen-type accessibility (popular _competitive_ and _unconstrained_ measure). What makes spatial availability a _constrained_ measure are the balancing factors that allocate opportunities proportionally. The balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).

```{r, synthetic-example-plot, fig.cap="\\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (x mode and z mode).", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/Fig1.png")
```

```{r access-calcs-table, include=FALSE}
#table with all the accessibility calculations
# Impedance parameter
b <- 0.1

# Calculate impedance function
od_tt_m <- od_tt_m %>%
  mutate(f = exp(-b * tt))
```

```{r calculating-beta-0.1-Vi-table-creation}
od_tt_m <- od_tt_m %>%
  mutate(catch = 1) %>%
  sp_avail_detailed(o_id = i, #spatial availability
                    d_id = j, 
                    pop = pop,
                    opp = opp,
                    r = catch,
                    f = f, 
                    alpha = 1)

V_i <- od_tt_m %>%
  group_by(i) %>%
  summarize(V_i = sum(V_ij))

od_tt_m <- od_tt_m %>%
  left_join(V_i, by = "i") %>%
  transmute(i,
            j,
            m,
            pop,
            opp,
            tt,
            V_ij = ifelse(V_ij < 0.001, "< 0.001", round(V_ij, 3)),
            V_i = ifelse(V_i < 0.001, "< 0.001", round(V_i, 3)),
            f = ifelse(f < 0.001, "< 0.001", round(f, 3)),
            f_p = ifelse(f_p < 0.001, "< 0.001", round(f_p, 3)),
            f_c = ifelse(f_c < 0.001, "< 0.001", round(f_c, 3)),
            f_t = ifelse(f_t < 0.001, "< 0.001", round(f_t, 3)))
```

```{r merging-data-for-table}
colnames(od_tt_m) <- c("i", "j","Mode", "Pop.", "Opp.", "TT", "V", "V_i", 
                       "f(TT)", "F^p", "F^c", "F^t")

od_tt_m_hux <- od_tt_m %>%
  as_hux()
```

```{r toy-example-table-all-values, cache=FALSE}
table_toy  <- od_tt_m_hux %>%
  set_background_color(everywhere, 7, "grey90") %>%
  set_background_color(everywhere, 8, "lightyellow") %>%
  set_bold(1, everywhere) %>%
  set_bold(everywhere, 1) %>%
  # set_bottom_border(c(1,2,5,10,15,20), everywhere) %>%
  # set_right_border(everywhere, c(1,4,7,10), brdr(1, "solid", "grey")) %>%
  # set_align(1:20, 1, "left") %>%
  # set_align(1:20,  2:10, "center") %>%
  # set_number_format(c(6,10,11:15,16:18), everywhere, 3) %>% #the smaller numbers, have 3 decimals
  # set_number_format(c(7:9,19,20), everywhere, 1) %>% # the larger numbers, have 1 decimal
  merge_cells(2:7, 1) %>% merge_cells(8:13, 1) %>% merge_cells(14:19, 1) %>%
  merge_cells(14:15, 2) %>% merge_cells(16:17, 2) %>% merge_cells(18:19, 2) %>% 
  merge_cells(2:4, 3) %>% merge_cells(5:7, 3) %>% merge_cells(8:10, 3) %>% 
  merge_cells(11:13, 3) %>% merge_cells(14:16, 3) %>% merge_cells(17:19, 3) %>% 
  merge_cells(2:4, 4) %>% merge_cells(5:7, 4) %>% merge_cells(8:10, 4) %>% 
  merge_cells(11:13, 4) %>% merge_cells(14:16, 4) %>% merge_cells(17:19, 4) %>% 
  merge_cells(2:7, 5) %>% merge_cells(8:13, 5) %>% merge_cells(14:19, 5) %>%
  merge_cells(2:7, 8) %>% merge_cells(8:13, 8) %>% merge_cells(14:19, 8) %>%
  set_caption("Summary description of the synthetic example: Spatial availability at each origin per mode $V_{im}$ (grey) and at each origin $V_i$ (yellow)") %>%
  set_label("tab:synthetic-example") %>% 
  set_latex_float("h!") %>% 
  set_width(1.1)

font_size(table_toy) <- 7

table_toy[1:19,]
```

We adapt the illustrative example used in @shen1998 assuming two modes and demonstrate spatial availability's within this multi-modal application. See the Figure \ref{fig:Fig1} and the Table \ref{tab:synthetic-example} with calculated spatial availability values (yellow - per origin and mode, and grey per origin). We use a negative exponential impedance function with $\beta=0.1$ as also used in @shen1998: we assume this impedance function for both modes.
$$
f(c_{ij}) = \exp(-\beta\cdot c_{ij})
$$

@@

According to the gravity modelling framework, the potential for interaction depends on the mass (i.e., the population) and the friction of distance (i.e., the impedance function). We begin by describing the proposed proportional allocation mechanism based on demand by the population. Recall, the total population in the example is 210,000. The proportion of the population by population center, by the two modes ($x$, and $z$) is:
$$
\begin{array}{l}
F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m=1}^M\sum_{i=1}^N P_{i}^m}\\

F^{pm}_A = \frac{50,000}{210,000} & \text{where } F^{px}_A = \frac{40,000}{210,000} \text{ & } F^{pz}_A = \frac{10,000}{210,000}\\ 
\\
F^{pm}_B = \frac{150,000}{210,000} & \text{where } F^{px}_B = \frac{100,000}{210,000} \text{ & } F^{pz}_B = \frac{50,000}{210,000}\\
\\
F^{pm}_C = \frac{10,000}{210,000} & \text{where } F^{px}_C = \frac{7,000}{210,000} \text{ & } F^{pz}_C = \frac{3,000}{210,000}\\
\end{array}
$$

Jobs are allocated proportionally from each employment center to each population center depending on their population sizes as per the balancing factors $F^p_i$. In this way, employment center 1 allocates $100,000\cdot \frac{50,000}{210,000}= 23,809.52$ jobs to $A$ (the suburb). Similarly, employment centers 2 and 3 allocate $100,000\cdot \frac{150,000}{210,000}= 71,428.57$ jobs to $B$ (the urban city) and $100,000\cdot \frac{10,000}{210,000}= 7,142.857$ jobs to $C$ (the satellite city). Notice how this mechanism ensures that the total number of jobs at employment center 1 is preserved at 100,000. The jobs at the employment centers are allocated proportional to the population size of the population center relative to all the population center sizes. Larger population centers are more competitive (i.e., they demand more) than smaller population centers, so they get allocated more job opportunities.

However, how do the populations access these jobs? In this synthetic example, the population either uses the regular mode $z$ or uses the faster mode $x$. Their choice in mode is already assigned: the mode selected can be seen as a characteristic of the population, or a way to subset the total population. In this way, the introduction of this population-characteristic of mode selection adds another layer of competition to the mass effect. We demonstrated that employment center 1 allocates $23,809.52$ jobs to $A$ but 20% of these jobs are allocated to population using $z$ ($10,000$ of the $50,000$ population in $A$) and 80% is allocated to the population using the faster mode $x$ ($40,000$ of the $50,000$ population in $A$). Similarly, employment centers 2 and 3 allocate $71,428.57$ jobs to $B$ (the urban city) of which 66% are allocated to the population using $z$ and 33% to the population using $x$ mode and the $7,142.857$ jobs allocated to $C$ (the satellite city) is split between 30% of the population using mode $z$ and 70% of the population using mode $x$.

Considering the subsets of the population that use the two different modes, the total number of jobs that are assigned to each population center and then to each sub-set of population, is still preserved. The total sum of all the jobs allocated in the region is still 210,000 in our example.

We can verify that the number of jobs allocated is consistent with the total number of jobs in the system:
$$
\begin{array}{l}
\text{Employment center 1 to population centers A, B, and C: }\\
100,000 \cdot \frac{40,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} + 100,000 \cdot \frac{100,000}{210,000} + 100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{7,000}{210,000} + 100,000 \cdot \frac{3,000}{210,000} = 100,000\\
\\
\text{Employment center 2 to population centers A, B, and C: }\\
100,000 \cdot \frac{40,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} + 100,000 \cdot \frac{100,000}{210,000} + 100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{7,000}{210,000} + 100,000 \cdot \frac{3,000}{210,000} = 100,000\\
\\
\text{Employment center 3 to population centers A, B, and C: }\\
10,000 \cdot \frac{40,000}{210,000} + 10,000 \cdot \frac{10,000}{210,000} + 10,000 \cdot \frac{100,000}{210,000} + 10,000 \cdot \frac{50,000}{210,000} + 10,000 \cdot \frac{7,000}{210,000} + 10,000 \cdot \frac{3,000}{210,000} = 10,000\\
\end{array}
$$

In the general case where there are $N$ population centers in the region, we define the following population-based balancing factors in Equation (\ref{eq:population-balancing-factor}):
```{=tex}
\begin{equation}
\label{eq:population-balancing-factor}
F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m=1}^M\sum_{i=1}^N P_{i}^m}
\end{equation}
```

Balancing factor $F^p_{i}$ corresponds to the proportion of the population in origin $i$ relative to the population in the region. On the right hand side of the equation, the numerator $P_{i}^m$ is the population at origin $i$ that use each type of mode $m$ for the full set of modes $M$.

The discussion so far concerns only the mass effect (i.e., population size) of the gravity model. In addition, the potential for interaction with opportunities is thought to increase with decreasing cost, so next we define similar balancing factors but based on the impedance.

@@

Clearly, exclusively using the population balancing factor $F^{pm}_{i}$ to calculate spatial availability does not account for the cost of reaching employment centers. Consider instead a set of balancing factors $F^{cm}_{ij}$ that accounts for the friction of distance. For this example, we define an impedance function $f(c_{ij}^m)$ that equals $\exp(-\beta\cdot c_{ij}^m)$ where $\beta = 0.1$ and the travel time $c_{ij}^m$ is either 15, 30 or 100 minutes for $m=z$ or 10, 20, or 80 minutes for $m=x$.

To illustrate, the impedance-based balancing factors $F^{cm}_{ij}$ would be the following for employment center 1 (employment center 2 and 3 have their own balancing factor values for each origin $i$ as is discussed later):
$$
\begin{array}{l}
F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m=1}^M \sum_{i=1}^N f(c_{ij}^m)}\\

F^{cx}_{A1} = \frac{0.223130}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.287349 \\
F^{cx}_{B1} = \frac{0.049787}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.06411619\\
F^{cx}_{C1} = \frac{0.000045}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.00005795144\\

F^{cz}_{A1} = \frac{0.3678794}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.4737587\\
F^{cz}_{B1} = \frac{0.1353353}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.1742861\\
F^{cz}_{C1} = \frac{0.0003354626}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.000432012\\
\end{array}
$$

Balancing factors $F^{cm}_{ij}$ use the impedance function to proportionally allocate more jobs to closer population centers, that is, to those with populations _more willing to reach the jobs_. Indeed, the factors $F^{cm}_{ij}$ can be thought of as the proportion of the population at $i$ willing to travel to destination $j$, conditional on the travel behavior as given by the impedance function. For instance, ${28.7349}\%$ of jobs from employment center 1 are allocated to population $A$ using the slower mode $x$ and ${47.37587}\%$ of the jobs are allocated to the population using the faster mode $z$ based on impedance. Population center $A$ is assigned $47.37587\%+{28.7349}\% = 76.11077\%$ of the jobs in the region based on the impedance, but it can be seen that the population using the faster mode is assigned a larger proportion of jobs. The population that uses the faster mode $z$ is more competitive than the population that uses the slower mode $x$.
